\textbf{Effective application of machine learning to microbiome based classification problems}

### 2019-10-09 submission/manuscript.tex
edits before sending to Jenna for 2nd time

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective application of machine learning to bacterial 16S rRNA
gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual relies on a rigorously
validated model. However, there are common methodological and reporting
problems that arise when applying ML to such data, that need to be
addressed for the field to progress. These problems include a lack of
transparency in which methods are used and how these methods are
implemented; evaluating models without a separate held-out test data;
large variation between the predictive performance on different folds of
cross-validation; and large variation between cross-validation and
testing performances. Nevertheless, the microbiome field is making
progress to avoid some of these pitfalls including validating their
models on independent datasets (7, 18, 19) and introducing ways to
better use ML tools (20--23). More work is needed to further improve
reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modelling approach has been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
models such as random forest and deep neural networks (2, 11, 24--26)
over simpler models such as logistic regression or other linear models
(18, 22, 27). Although complex models may in some cases, capture
important non-linear relationships and therefore yield better
predictions, they may result in black boxes that lack interpretibility.
Such models require post hoc explanations to quantify the importance of
each feature in making predictions. Depending on the goal of the
modeling, researchers may choose to use different approaches. For
example, researchers trying to identify the populations of microbiota
associated wwith disease may desire a more interpretable model, whereas
clinicians may emphasize predictive performance. Nonetheless, it is
important to understand the tradeoff and accepting that this tradeoff
may be minimal (28, 29). It is important for researchers to justify
their choice of modelling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and training time. To
demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

Model selection required tuning hyperparameters. Hyperparameters are
parameters which need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, (C) is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0, where the model's predictions are perfectly
incorrect. An AUROC value of 0.5 indicates that model's predictions are
no differect than random. To select hyperparameters, we performed a full
grid search for hyperparameter settings when training our models.
Default hyperparameter settings in previously developed ML packages in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization strengths (C).

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had higher test AUROC values than the other models
for detecting SRNs (Resampling test from two groups, p \textless{}
0.05). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and
L2-regularized SVM with linear kernel) performed nearly as well
non-linear models. To support our claim, we investigated the
distribution of differences between random forest and L2-regularized
logistic regression AUROC values for each data-split (Figure S3). Our
results underscored that L2-regularized logistic regression perform
nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
difference in median AUROCs was 0.021 in L1-regularized SVM with linear
kernel, followed by SVM with radial basis function kernel and decision
tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
These differences are relatively small and gives us confidence in our
estimate of the generalization performance of the models.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
variation in cross-validation performance within both models at lower
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (31--33). Because we often use ML models not just to predict
a health outcome but also to identify the biomarkers for a disease,
model interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTU 50,
426, 609, 822, 1239). The benefit of this approach was that the results
of the analysis were based on the trained model parameters and provided
information regarding the sign and magnitude of the impact of each OTU.
However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model, in which we randomly permuted
groups of perfectly correlated features together and other features
individually across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e,
testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTU 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTU
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
In this case, similar to previous studies (22, 34), it was not possible
to distinguish between health and disease using a single OTU. Although
permutation analysis allowed us to gauge the importance of an OTU, the
analysis was post-hoc (i.e.~done using the test data) and these results
did not allow us to directly interrogate the models to know whether an
OTU was associated with classifying a subject as being healthy or having
an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S4{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. Although permutation analysis does not not allow
us to determine the weight or the sign of the features, these results
indicate that both methods are consistent in selecting the most
important OTUs.

@@

To address these needs, we developed an open-sourced framework to ML
models. using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference of only 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model and
is most often explained using post-hoc methods such as permutation
importance. On the other hand, using L2-regularized logistic regression
we ranked the importance of each OTU based on their feature weights.
Comparing many different models showed us that the most complex model
was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that predicts a binary health
outcome. We performed a random data-split to create a training set (80\%
of the data) and a held-out test set (20\% of the data), which we used
to evaluate predictive performance. We repeated this data-split 100
times to measure the possible variation in predictive performance.
During the training, we tuned the model hyperparameters with a repeated
five-fold cross-validation. Despite the high number of features
microbiome datasets typically have, the models we built with this
pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (35). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we took collinearity into consideration by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other however, we can reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to do
correlation analyses and further experimentation for biological
validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as number of classes increases, more samples are required
for each category to train an accurate model. We also did not use
regression-based analyses to predict a non-categorical outcome. We have
previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (36). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (11, 37, 38) but microbiome datasets often
suffer from having many features and small sample sizes, which can
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly harder and distinct undertaking. For
example, we currently lack standardized methods to collect patient
samples, generate sequence data, and report clinical data. We are also
challenged by the practical constraints of OTU-based approaches. The de
novo algorithms commonly in use are slow, require considerable memory,
and result in different OTU assignments as new data are added. Finally,
we also need independent validation cohorts to test the performance of a
diagnostic model. To realize the potential for using ML approaches with
microbiome data, it is necessary that we direct our efforts to overcome
these challenges.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A full
grid search was needed to avoid large variation in prediction
performance. For L2-regularized logistic regression, L1 and
L2-regularized SVM with linear and radial basis function kernels, we
tuned the \textbf{cost} hyperparameter which determines the
regularization strength where smaller values specify stronger
regularization. For SVM with radial basis function kernel we also tuned
\textbf{sigma} hyperparameter which determines the reach of a single
training instance where for a high value of sigma, the SVM decision
boundary will be dependent on the points that are closest to the
decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For XGBoost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. Recently developed tools such
as Hyperband help researchers with hyperparameter selection that can be
incorporated to microbiome studies (42) .

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to microbiome based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective application of machine learning to microbiome based
classification problems}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population seperately (1). Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because we are able to sample large numbers of individuals
and because the models are effective at accounting for the interpersonal
microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; large variation between
the predictive performance on different folds of cross-validation; and
large variation between cross-validation and testing performances.
Nevertheless, the microbiome field is making progress to avoid some of
these pitfalls including validating their models on independent datasets
(8, 19, 20) and introducing ways to better use ML tools (21--24). More
work is needed to further improve reproducibility and minimize
overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
models such as random forest and deep neural networks (3, 12, 25--27)
over simpler models such as logistic regression or other linear models
(19, 23, 28). Although in some cases complex models may capture
important non-linear relationships and therefore yield better
predictions, they result in black boxes that lack interpretability. Such
models require post hoc explanations to quantify the importance of each
feature in making predictions. Depending on the goal of the modeling,
researchers may choose to use different approaches. For example,
researchers trying to identify the microbiota associated with disease
may desire a more interpretable model, whereas clinicians may emphasize
predictive performance. Nonetheless, it is important to understand the
tradeoff and accept that the benefit of more complex, less interpretable
models may be minimal (29, 30). It is important for researchers to
justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and training time. To
demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classify individuals as having normal colons or colonic lesions based on
the 16S rRNA gene sequences collected from fecal samples (4). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset we developed a ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

Model selection required tuning hyperparameters. Hyperparameters are
parameters which need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that model's predictions are no
different than random. To select hyperparameters, we performed a full
grid search for hyperparameter settings when training the models.
Default hyperparameter settings in previously developed ML packages
available in R, Python, and MATLAB programming languages may be
inadequate for effective application of classification algorithms and
need to be optimized for each new ML task. For example, L1-regularized
SVM with linear kernel showed large variability between different
regularization strengths (C) and needed to be tuned to find the optimal
C value {[}Figure S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had higher test AUROC values than the other models
for detecting SRNs (resampling test between means of two groups, p
\textless{} 0.01). The median AUROC of the random forest model was 0.695
{[}IQR 0.650-0.739{]}. L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models. To support our
claim, we investigated the distribution of differences between random
forest and L2-regularized logistic regression AUROC values for each
data-split (Figure S3). Our results underscored that L2-regularized
logistic regression perform nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
difference in median AUROCs was 0.021 in L1-regularized SVM with linear
kernel, followed by SVM with radial basis function kernel and decision
tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
These differences are relatively small and give us confidence in our
estimate of the generalization performance of the models.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
variation in cross-validation performance within both models at lower
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (32--34). Because we often use ML models not just to predict
a health outcome but also to identify the biomarkers for a disease,
model interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a post hoc explanation of the model, in which we randomly permuted
groups of perfectly correlated features together and other features
individually across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e,
testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
In this case, similar to previous studies (23, 35), it was not possible
to distinguish between health and disease using a single OTU. Although
permutation analysis allowed us to gauge the importance of an OTU, the
analysis was post hoc (i.e.~done using the test data) and these results
did not allow us to directly interrogate the models to know whether an
OTU was associated with classifying a subject as being healthy or having
an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. Although permutation analysis does not not allow
us to determine the weight or the sign of the features, these results
indicate that both methods are consistent in selecting the most
important OTUs.

@@

To address these needs, we developed an open-sourced framework to ML
models. Using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference less than 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model and
is most often explained using post hoc methods such as permutation
importance. On the other hand, using L2-regularized logistic regression
we ranked the importance of each OTU based on their feature weights.
Comparing many different models showed us that the most complex model
was not necessarily the best model for our ML task.

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
repeated this data-split 100 times to measure the possible variation in
predictive performance. During the training, we tuned the model
hyperparameters with a repeated five-fold cross-validation. Despite the
high number of features microbiome datasets typically have, the models
we built with this pipeline were generalizable as shown by the good test
AUROCs.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (36). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we took collinearity into consideration by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other, however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to do
correlation analyses and further experimentation for biological
validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (37). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 38, 39) but microbiome datasets often
suffer from having many features and small sample sizes, which result in
overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly harder and distinct undertaking. For
example, we currently lack standardized methods to collect patient
samples, generate sequence data, and report clinical data. We are also
challenged by the practical constraints of OTU-based approaches. The de
novo algorithms commonly in use are slow, require considerable memory,
and result in different OTU assignments as new data are added (40).
Finally, we also need independent validation cohorts to test the
performance of a diagnostic model. To realize the potential for using ML
approaches with microbiome data, it is necessary that we direct our
efforts to overcome these challenges.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A full
grid search was needed to avoid large variation in prediction
performance. For L2-regularized logistic regression, L1 and
L2-regularized SVM with linear and radial basis function kernels, we
tuned the cost hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned sigma hyperparameter
which determines the reach of a single training instance where for a
high value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where deeper the tree, the more
splits it has. For random forest, we tuned the number of features to
consider when looking for the best tree split. For XGBoost, we tuned for
learning rate and the fraction of samples to be used for fitting the
individual base learners. Recently developed tools such as Hyperband
help researchers with hyperparameter selection that can be incorporated
to microbiome studies (44) .
